---
title: "Multiple linear regression"
author: 'Samuel Iguku Kigamba'
date: '11/28/2019'
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment?

        This is an observational study.
        
   The original research question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.
    
    Observational study cannot show caustion, only association.
    rephrased question: Is there a relationship between beauty and differences in course evaluations?

2.  Describe the distribution of `score`. Is the distribution skewed? 
        
        Based on the histogram, the distribution of score is unimodal and left-skewed, centered around 4.5.
        
What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not?

        Most students give high marks to courses with courses with fewer students giving low marks.
    
```{r}

hist(evals$score)

```

         
        

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).
    
        Variables: age and language.
        Lets use side by side box plots for the visualization.
        The mean age for two categories is very similar and IQR ranges are almost the same. 
        The age of English-speaking professor reaches higher than the non-English speaking professors,
```{r}

plot(evals$age ~ evals$language, 
     xlab = "Lang", ylab = "Age")

```

```{r}

table(evals$language)

```

        Note the huge variance between english and non english professors.


## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg, eval = TRUE}
plot(evals$score ~ evals$bty_avg)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

        There are 463 observations in the data frame but there are points less than
        the number of observations in the scatterplot. There could be overlapping points.

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?
    
```{r}

plot(jitter(evals$score) ~ evals$bty_avg)

```

        Using jitters we are able to view overlapping points, those with the same bty_avg.
        

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
```{r}

m_bty <- lm(evals$score ~ evals$bty_avg)
plot(jitter(evals$score) ~ evals$bty_avg)
abline(m_bty)

```

```{r}

summary(m_bty)

```

        Linear model equation: 
\[score^ = 3.88034 + 0.06664 × avg beauty rating\]

        Interpretation: For every point in the average beauty rating, the score increases by 0.067. 
        The p???value from the summary above is close to zero, which means that even at a significance level of 0.001, 
        the slope is not due to chance and there is a statistically significant relationship. 
        

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).
    
```{r}

plot(m_bty$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)

```

        The relationship appears to be linear since there is no apparent pattern in the residuals plot.
        There is constant variability of points.
        
```{r}
hist(m_bty$residuals)

```

```{r}

qqnorm(m_bty$residuals)
qqline(m_bty$residuals)

```

        The histogram of residuals is left-skewed.
        The normal probability plot deviates from the diagonal line on both ends.
        The residuals are not normally distributed from this observations.

        Assumptions:
        Each student acted independently. 
        Some students might have been influenced by their fellow students.


## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel, eval = TRUE}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels, eval = TRUE}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender, eval = TRUE}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.
    
```{r}

qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals)

```
    
        Residuals deviate significantly from the normal probability line and the distribution seems
        to be left-skewed. 
        The residuals of the model are not nearly normal. 
        Assumption: The distribution of the variables is near normal.
        
```{r}

plot(abs(m_bty_gen$residuals) ~ m_bty_gen$fitted.values,
     xlab = "Fitted Values", ylab = "Absolute Value of Residuals")

```

        There is limited variability of the residuals on both ends. Given the left-skew of the residuals, 
        there is doubt as to the mear constant variability.

        Assumption: independence of observations.
        
```{r}
plot(m_bty_gen$residuals ~ evals$gender)

```
        Variability does not change much with the gender variable.

        The residuals are not near normal and variability is not necessarily nearly constant. 
        Care needs to be taken in reporting the results of this model.
        
        
8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
        The parameter estimate for bty_avg has changed with addition of the gender variable, but it remains 
        a significant predictor of score.

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines, eval = TRUE}
multiLines(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?
    
        Equation of the line corresponding to males:
\[score^ = ??^0 + ??^1 × bty avg + ??^2\]

        Evaluation score from males is higher.

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.
    
```{r}

m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)

```
        
The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.
    
    cls_level have the highest p-value in this model because class level will not
    have any association with the professor score..

Let's run the model...

```{r m_full, eval = TRUE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.
    
    Although class level has the second highest p-value, the variable class professor:
    number of professors teaching sections in course in sample, has the highest p-value.

13. Interpret the coefficient associated with the ethnicity variable.

        The slope of ethnicity variable is 0.1234929. 
        The standard error for the slope is 0.0786273. 
        The t-test statistic for the null hypothesis is 1.571. 
        The p-value for a two-sided alternative hypothes is: 0.11698.

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?
    
```{r}

m_no_profs <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_no_profs)

```
        
        After dropping cls_profs, the coefficients and significance of the other explanatory 
        variables did not change.

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.
    
    Based on the significance level of 0.05, the best model is below.

````{r}

m_best <- lm(score ~ ethnicity + gender + language + age + cls_perc_eval 
             + cls_credits + bty_avg + pic_color, data = evals)
summary(m_best)

```
\[score^ = 3.77 + 0.17 × ethnicity + 0.21 × gender minus0.21 × language minus0.01 × age + 0.005 × cls perc eval + 0.51 × cls credits + 0.05 × bty avg minus0.19 × pic color\]


16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
    
```{r}

qqnorm(m_best$residuals)
qqline(m_best$residuals)

```

```{r}

plot(abs(m_bty_gen$residuals) ~ m_bty_gen$fitted.values,
     xlab = "Fitted Values", ylab = "Absolute Value of Residuals")

```

        The residuals of the model are nearly normal as shown in the QQ plot. 
        While there are a few observations that deviate noticeably from the line, they are not particularly extreme.
        
17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?
    
    This new information will influence the assumptions concerning independence of observations. 
    However it is not expected to influence the conditions.

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
    Based on the model, the highest score will be associated with professors who:

        is from an English speaking school, has a black & white picture, not part of a minority group,
        on a younger side, teaches one credit courses, has a high beauty ranking and is male.


19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
    
    I would not generalize my conclusiona to apply to professors at other university because this is an observational study
    and  the sample size is too small(6 students). In addtion, the rating is subjective and the judgement on beauty could 
    be quite different from one school to the other school due to the different culture.
    
    

